A demonstration of the \LaTeXe\ class file for \itshape{SAGE Publications}

### 2015-12-22 body/equilibrium.tex
andrea's modifications

% !TEX root =  ../main.tex
\section{Heuristics for relevant contact selection}

\subsection{A heuristic for robust static equilibrium}
The planner is designed so that any generated contact configuration is in static equilibrium.
Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real-world application, regardless of perception and control uncertainties.

We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness\deleted{for any configuration}. While the presented LP is original, it is based on an analysis of the problem that we proposed in \cite{Prete2016}, where the interested reader can find more details.

@@

According to Coulomb's law, \replaced{slipping occurs only if a contact force is on the boundaries of the friction cone.}{ the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.}
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be expressed as a convex combination of its four generating rays.
\deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
\end{equation*}
\added{where $\bm{\beta}_i$ contains the coefficient of the cone generators.}
We can then stack all the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta},
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = (\mathbf{f}_0,...,\mathbf{f}_e)$.

@@

%~ Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul2d}, it means that the configuration is statically balanced.
The problem can then be formulated as an LP:

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

@@

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \bm{1}\\
\end{aligned} \end{equation}

@@

In our implementation, rather than solving directly \eqref{eq:lin_prog_rob}, we solve an equivalent problem of smaller dimension that we get by taking the dual of \eqref{eq:lin_prog_rob} and eliminating the Lagrange multipliers associated to the inequality constraints:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

>>>>

% !TEX root =  ../main.tex
\section{Heuristics for relevant contact selection}
\label{sec:heuristics}
\subsection{A heuristic for robust static equilibrium}
The planner is designed so that any generated contact configuration is in static equilibrium.
Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real-world application, regardless of perception and control uncertainties.

We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness. While the presented LP is original, it is based on an analysis of the problem that we proposed in \citep{Prete2016}, where the interested reader can find more details.

@@

According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be expressed as a convex combination of its four generating rays.
%~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
\end{equation*}
where $\bm{\beta}_i$ contains the coefficient of the cone generators.
We can then stack all the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta},
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = (\mathbf{f}_0,...,\mathbf{f}_e)$.

@@

%~ Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
%~ The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
%~ \begin{align} \label{eq:new_eul2d}
%~ \mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
%~ \end{align}

If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul}, it means that the configuration is statically balanced.
The problem can then be formulated as a LP:

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e} \\
%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

@@

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D} \mathbf{c} + \mathbf{d} \\
%~ \st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \bm{1}\\
\end{aligned} \end{equation}

@@

In our implementation, rather than solving directly \eqref{eq:lin_prog_rob}, we solve an equivalent problem of smaller dimension that we get by taking the dual of \eqref{eq:lin_prog_rob} and eliminating the Lagrange multipliers associated to the inequality constraints:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D} \mathbf{c} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

### 2015-12-22 body/intro.tex
andrea's modifications

% !TEX root =  ../main.tex
We consider the problem of planning the \adnote{an} acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment. Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support \adnote{maintain} the equilibrium of the system.

Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. But \adnote{don't start sentences with 'but', you can start with 'however' followed by a comma} executing
this behavior on cluttered environments is dangerous, if not impossible.
In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
noted: ``Except for egress, no robots in the DRC Finals used
the  stair  railings  or  any  form  of  bracing.   Even  drunk  people  are  smart  enough  to  use  nearby  supports.
Full body locomotion (handholds,  bracing,  leaning against a wall or obstacles) should be easier than our
current high performance minimum contact locomotion approaches."

Indeed, the current approach to locomotion planning is to avoid obstacles as much as possible, instead of using them
to facilitate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
mechanically easier, but algorithmically more complicated for planning, and the transitions are difficult to
both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of valid \adnote{'valid' is not super clear to me here, I'd rather say 'collision-free'} configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.

Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
More recent successes use a local solution, resulting in more reasonable performances \adnote{'performances' is ambiguous, I'd rather use 'computation times'} (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
its execution is too costly.

As suggested by \citeauthor{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot trajectory.
This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly, \adnote{I don't like having two adverbs ending in 'ly' one after the other, 'extremely fast' sounds better to me}
while preserving a success rate greater than TODO \% in the demonstrated scenarios.

In the remainder of this introduction, we discuss further the current state of the art. This allows us to then \adnote{this 'then' sounds bad here} situate more precisely our contributions.

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches can intrinsically not \adnote{can not intrinsically? are intrinsically not able to?} adapt to new situations or discover complex behaviors in unforeseen contexts.

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: \adnote{planning? finding? computing?} a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial \adnote{http://www.thefreedictionary.com/nontrivial} problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable \adnote{'scalable' means 'capable of being scaled'} to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, at the cost of prohibitive computation times (several hours). Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance \adnote{I don't like 'relevance', what about 'quality'?} of the guide trajectory,  which is not evaluated \textit{a priori}. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium. Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented). Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.

It thus appears that an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms
existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-executable \adnote{dynamically-feasible?}
trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances.
To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 
Such a planner holds the promise of a complete real-time multi-contact locomotion system.

\subsection{Paper contribution and organization}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the \adnote{generation of the} guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}. We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
\item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
\end{itemize}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
For each problem, we introduce an original formulation of the related theoretical issues.
Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenario, or even better, for N robots and M scenarios}. The complete framework is implemented within
the open source platform HPP.

@@

To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
condition can be verified in a low-dimensional space, that we call $C_{reach}$. This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained. This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.

To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain extremely rapidly \adnote{again, extremely rapidly} a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics.
In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
the Extended FORce Transmission ratio (EFORT).

@@

In comparison with previous work, \adnote{Compared to previous work/Contrary to previous work [cite Mordatch,...]} our planner does not produce a complete motion, but a discrete sequence of contacts.
Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
is the fastest multi-contact planning solution to our knowledge.

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with a \adnote{I remove the word 'argumented' because it does not exist and I've not been able to find a good replacement} discussion on the implementation of this theory to real-world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. TODO Appendix describes the generation process
for the robot.

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:efort} present respectively our answer to \adnote{I remove 'problems'} \Pa and \Pb. In Section~\ref{heuristics}, we present two heuristics used for the selection of \adnote{I remove 'the best'} a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.

>>>>

% !TEX root =  ../main.tex
We consider the problem of planning an acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment. Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to maintain the equilibrium of the system.

Most multipedal locomotion systems focus on cyclic walking gaits \citep{Kajita03a}. However, executing
this behavior on cluttered environments is dangerous, if not impossible.
In an analysis of their participation to the Darpa Robotics Challenge, \citeauthor{atkensondarpa}
noted: ``Except for egress, no robots in the DRC Finals used
the  stair  railings  or  any  form  of  bracing.   Even  drunk  people  are  smart  enough  to  use  nearby  supports.
Full body locomotion (handholds,  bracing,  leaning against a wall or obstacles) should be easier than our
current high performance minimum contact locomotion approaches."

Indeed, the current approach to locomotion planning is to avoid obstacles as much as possible, instead of using them
to facilitate locomotion. The reason for this, as the authors state, is that ``More contacts make tasks
mechanically easier, but algorithmically more complicated for planning, and the transitions are difficult to
both plan and control[...].  We have seen very few robot planners that are  capable of  generating this  behavior."
The difficulty of addressing such a problem comes both in practice from the proximity to the obstacles (that tends to make the sampling of collision-free configuration tedious) and in theory from the foliation of the configuration space, where zero-measure manifolds intersect in a combinatorial manner \citep{simeon-manipulation-04}.

Previous contributions that embrace the combinatorial provide complete approaches, not applicable in practice because they require hours of computations \citep{conf/iser/BretlRLKA04}.
More recent successes use a local solution, resulting in more reasonable computation times (still far from real time), at the cost of dynamically inconsistent behaviors \citep{Mordatch:2012:DCB:2185520.2185539}.
Neither global nor local methods present satisfying performances, because planning simultaneously the robot trajectory and the contacts that allow
its execution is too costly.

As suggested by \citeauthor{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot trajectory.
This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely fast.
while preserving a success rate greater than TODO \% in the demonstrated scenarios.

In the remainder of this introduction, we discuss further the current state of the art. This allows us to situate more precisely our contributions.

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are intrinsically not able to adapt to new situations or discover complex behaviors in unforeseen contexts.

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, at the cost of prohibitive computation times (several hours). Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the quality of the guide trajectory,  which is not evaluated \textit{a priori}. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium. Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented). Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.

It thus appears that an integrated optimization-based approach \citep{Mordatch:2012:DCB:2185520.2185539}, able to generate a locally-optimal, complete trajectory within minutes, currently outperforms
existing planners, decoupled or not. However, recent contributions are able to interpolate dynamically-feasible
trajectories between contact configurations \citep{herzog2015trajectory, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with real-time performances.
To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. 
Such a planner holds the promise of a complete real-time multi-contact locomotion system.

\subsection{Paper contribution and organization}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces). We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
\item There are infinite combinations of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions.
\end{itemize}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
For each problem, we introduce an original formulation of the related theoretical issues.
Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
The relaxation is justified empirically, with statistics presented for 2 robots and 4 scenarios. The complete framework is implemented within
the open-source platform HPP.

@@

To address \Pa we rigorously formulate a geometric condition that approximates the true feasibility of a path, the reachability condition. We claim that this 
condition can be verified in a low-dimensional space, which we call $C_{reach}$. This formulation can make the planning of a guide path more efficient computationally, while providing similar guarantees to planning directly in the configuration space.
In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained. This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot in a timespan ranging from a few milliseconds to a few seconds, depending on the scenario.

To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain rapidly a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, as well as a set of user-defined heuristics.
In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
the Extended FORce Transmission ratio (EFORT).

@@

Compared to previous work \citep{Mordatch:2012:DCB:2185520.2185539} our planner does not produce a complete motion, but a discrete sequence of contacts.
Since the interpolation of a complete trajectory between such configurations can be solved within real-time constraints \citep{Carpentier2016}, we claim that our planner
is the fastest multi-contact planning solution to our knowledge.

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with a discussion on the implementation of this theory to real-world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. Appendix~\ref{app:rom} describes the generation process
for the robot.

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb. In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.
